---
params: 
  saved: none
  species: Test Species Name
  assessor_name: Dr. Curious
  geo_location: Earth
  include_about: true
  debug: true
title: "`r params$species`"
subtitle: Climate Change Vulnerability Index Report
abstract: "`r params$geo_location`"
abstract-title: "Geographic Location"
date: today
published-title: "Date"
author: "`r params$assessor_name`"
format: 
  html:
    page-layout: full
    theme: 
      - default
      - styles.scss
keep-md: false
fig-responsive: false
---
<!-- Notes 
fig-responsive: false (above) is required to keep the spacing around the 
plotly dials appropriate.

For debugging, `keep-md: true` and compile this script directly (don't use `build_report()`)
Then check inst/qmd/results_report.html.md for intermediate problems
-->


## Overview

```{r setup}
#| include: false

library(purrr)
library(dplyr)
library(ggplot2)
library(ccviR)
library(sf)
library(htmltools)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = FALSE)

saved <- params$saved
species <- params$species
include_about <- as.logical(params$include_about)
debug <- params$debug

# defined as latex colors below
#cols <- c("gray", "green", "yellow", "orange", "red")
index_labs <- tribble(
  ~code, ~label,                 ~col,      ~text_col,
  "IE", "Insufficient Evidence", "#808080", "black",
  "LV", "Less Vulnerable",       "#008000", "white",
  "MV", "Moderately Vulnerable", "#FFC125", "black",
  "HV", "Highly Vulnerable",     "#FF8C00", "black",
  "EV", "Extremely Vulnerable",  "#FF0000", "white",
  "NA", "Not Applicable",        "white",   "black")

value_labs <- index_labs %>%
  filter(code != "NA") %>%
  select(col, text_col) %>%
  mutate(value = c("Unknown", rev(ccviR:::valueNms)),
         opts = c(-1, rev(ccviR:::valueOpts)))


# Custom boxes - bootstrap 5
# cards 
#  - https://getbootstrap.com/docs/5.3/components/card/#horizontal
#  - https://getbootstrap.com/docs/5.3/components/card/#background-and-color
# Quarto uses new grid system https://getbootstrap.com/docs/5.3/layout/css-grid/
# mb-3 - https://getbootstrap.com/docs/5.3/utilities/spacing/#margin-and-padding
# 
# TODO: Not currently used
results_box <- function(s, column) {
  map(seq_len(nrow(s)), ~ {
    div(
      class = c("card mb-2", paste0("card-", tolower(s[[column]][.x]))),
      div(class="card-header", s[["scenario_name"]][[.x]]),
      div( 
        class = "card-body",
        h4(class = "card-title", 
           index_labs$label[index_labs$code == s[[column]][.x]]), 
      )
    )
  })
}

vuln_box <- function(tbl) {
  split(tbl, tbl$Code) %>%
    map(~{
      div(class = "card mb-3",   # mb-2 means Margin-Bottom 2 (can change 2)
          div(class = "card-header", 
              h4(strong(.x$Code[1], .noWS = "after"), ": ", .x$Question[1])),
          div(class = "card-body", 
              div(class = "card-title", 
                  map2(.x$scenario_name, .x$Value, 
                       ~ tagList(strong(.x), .y))),
              strong("Comments:"), .x$com[1], br(),
              strong("Evidence:"), .x$evi[1]
          )
      )
    })
}
```


```{r set_saved}
#| include: false
# If testing use test saved data
if(length(saved) == 1 && saved == "none") {
  saved <- ccviR:::test_files()$saved$full_run %>%
    ccviR:::load_previous()
} else {
  # Otherwise remember that Quarto isn't R! So can handle lists, but not data.frames
  # So convert lists back to df.
  saved <- as.data.frame(saved) %>%
    readr::type_convert(na = ".na")
  
  # Debugging: 
  if(debug) writeLines(capture.output(saved), "test.txt")
}
```

```{r load_spatial}
#| include: false
range_poly <- read_sf(saved$rng_poly_pth[1])
assess_poly <- read_sf(saved$assess_poly_pth[1])
clim_vars <- get_clim_vars(saved$clim_dir_pth[1], scenario_names = saved$scenario_name, quiet = TRUE)
clim_readme <- fs::path(saved$clim_dir_pth[1], "climate_data_readme.csv") %>%
  utils::read.csv(check.names = FALSE)
spat_tbl <- ccviR:::apply_spat_tholds(saved, saved$cave)
```




<!-- ::::{.grid} -->

<!-- :::{.g-col-6} -->

<!-- #### Climate Change Vulnerability Index -->

<!-- ```{r} -->
<!-- #| output: asis -->
<!-- # TODO: Keep these? Or just the dials? -->
<!-- b <- results_box(saved, "CCVI_index") -->
<!-- tagList(!!!b) -->
<!-- ``` -->
<!-- ::: -->

<!-- :::{.g-col-6} -->

<!-- #### Migratory Exposure Index -->

<!-- ```{r} -->
<!-- #| output: asis -->
<!-- b <- results_box(saved, "mig_exposure") -->
<!-- tagList(!!!b) -->
<!-- ``` -->
<!-- ::: -->

<!-- :::: -->


#### Climate Change Vulnerability Index
```{r}
dials_index <- purrr::map2(saved$CCVI_index, saved$scenario_name, ~ {
  ccviR:::plot_index_gauge(.x, type = "index", ttl = .y)
})
# break-inside etc. forces page breaks around the figures
# margin: auto centres the figures
div(style = "break-inside: avoid; break-after: auto",
    div(class = "grid", 
        !!!purrr::map(dials_index,
                      ~div(class = "g-col-6", .x, style = "margin: auto"))
    )
)
```

#### Migratory Exposure Index
```{r}
dials_mig_exp <- purrr::map2(saved$mig_exposure, saved$scenario_name, ~ {
  ccviR:::plot_index_gauge(.x, type = "mig_exp", ttl = .y)
})
# break-inside etc. forces page breaks around the figures
# margin: auto centres the figures
div(style = "break-inside: avoid; break-after: auto",
    div(class = "grid", 
        !!!purrr::map(dials_mig_exp, 
                      ~div(class = "g-col-6", .x, style = "margin: auto"))
    )
)
```

### Species range
```{r range_map}
#| fig-cap: "Map of the species range (red) and the assessment area (black)."

# TODO: should add a context polygon that is North America to state/prov level as
# background then choose the map extent as whichever is bigger between range and assessment area

comb_ext <- sf::st_union(sf::st_bbox(range_poly) %>% sf::st_as_sfc(),
                         sf::st_bbox(assess_poly) %>% sf::st_as_sfc() %>%
                         sf::st_transform(sf::st_crs(range_poly))) %>%
  sf::st_transform(4326)

ggplot() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  geom_sf(data = st_crop(context_North_Am, comb_ext), fill = "grey90", colour = "grey80") +
  geom_sf(data = assess_poly, aes(colour = "Assessment Area"), fill = NA, linewidth = 0.5, 
          show.legend = "line") +
  geom_sf(data = range_poly, aes(colour = "Range"), linewidth = 0.5, fill = NA, 
          show.legend = "line") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_colour_manual(name = NULL, values = c("Assessment Area" = "black", "Range" = "red"))
```

## Vulnerability Factors

### Section A. Exposure to Local Climate Change

```{r bivar_exp}
#| fig-height: 3
#| fig-width: 7
#| fig-cap: "Map of exposure based on both change in temperature and change in moisture regime."

bivar_exp_plot <- plot_bivar_exp(clim_vars$mat, clim_vars$cmd, assess_poly, range_poly)

gridExtra::grid.arrange(grobs = list(ggplot2::ggplotGrob(bivar_exp_plot$plot),
                                     ggplot2::ggplotGrob(bivar_exp_plot$legend)),
                        layout_matrix = rbind(c(1,1,1,1,1,2),
                                              c(1,1,1,1,1,2),
                                              c(1,1,1,1,1,2)))
```

#### Temperature
```{r temp_tbl}
#| output: "asis"
ccviR:::get_exposure_table(spat_tbl, "MAT", clim_readme, clim_readme$brks_mat)
```

#### Moisture
```{r moist_tbl}
#| output: "asis"
ccviR:::get_exposure_table(spat_tbl, "CMD", clim_readme, clim_readme$brks_cmd)
```

#### CCEI
```{r ccei_tbl}
#| output: "asis"
if(any(!is.na(select(saved, contains("CCEI_"))))) {
  ccviR:::get_exposure_table(spat_tbl, "CCEI", clim_readme, clim_readme$brks_ccei)
} else {
  div("No Migratory Exposure based on the Climate Change Exposure Index availble")
}
```



```{r vuln_tbl}
#| output: "asis"
coms <- saved %>%
  select(scenario_name, matches("(evi|com)_[B,C,D]\\d.*")) %>%
  tidyr::pivot_longer(-scenario_name, 
                      names_to = c(".value", "Code"),
                      names_pattern = "(evi|com)_(.+)") %>%
  mutate(com = as.character(com), evi = as.character(evi),
         com = tidyr::replace_na(com, ""),
         evi = tidyr::replace_na(evi, ""))

vuln_qs <- saved %>%
  select(scenario_name, matches("^[B,C,D]\\d.*")) %>%
  mutate(across(everything(), as.character)) %>%
  tidyr::pivot_longer(-scenario_name, names_to = "Code", values_to = "Score") %>%
  tidyr::separate(Score, into = paste0("Value", 1:4), fill = "right", sep = ", ",
                  convert = TRUE) %>%
  mutate(Value1 = tidyr::replace_na(Value1, -1)) %>%
  mutate(score = 1) 

vuln_qs <- vuln_qs %>%
  ccviR:::filter_applicable_qs(saved$tax_grp, 
                               c("Vascular Plant", "Nonvascular Plant")) %>%
  tidyr::pivot_longer(cols = contains("Value"), values_to = "opts") %>%
  left_join(value_labs, by = "opts") %>%
  mutate(value = factor(value, levels = value_labs$value)) %>%
  filter(opts == min(opts, na.rm = TRUE) | opts == max(opts, na.rm = TRUE),
         .by = c("scenario_name", "Code")) %>%
  filter(!is.na(score))

# Debugging
if(debug) writeLines(capture.output(vuln_qs), "test_vuln_intermediate.txt")

vuln_qs <- vuln_qs %>%
  mutate(
    Value = pmap(
      list(value, col, text_col),
      ~span(..1, style = paste("padding: 2px; font-size: 90%; ",
                               "background-color:", ..2, "; ",
                               "color:", ..3)))) %>%
  summarize(Value = tagList(Value), .by = c("scenario_name", "Code"))
#mutate(Value = stringr::str_replace_all(
#  Value, stats::setNames(value_labs$value, as.character(value_labs$opts))))


vuln_tbl <- left_join(vuln_qs, coms, by = join_by(scenario_name, Code)) %>%
  left_join(vulnq_code_lu_tbl, by = join_by(Code)) %>%
  select(scenario_name, Question, Code, Value, com, evi)
  #left_join(select(value_labs, -"opts"), by = c("Value" = "value"))

# Debugging
if(debug) writeLines(capture.output(vuln_tbl), "test_vuln.txt")
```

### Section B. Indirect Exposure to Climate Change
```{r section_b}
#| output: asis
b <- vuln_box(filter(vuln_tbl, stringr::str_detect(Code, "^B")))
tagList(!!!b)
```

### Section C. Sensitivity and Adaptive Capacity
```{r section_c}
#| output: asis
c <- vuln_box(filter(vuln_tbl, stringr::str_detect(Code, "^C")))
tagList(!!!c)
```

### Section D. Documented or Modeled Response to Climate Change
```{r section_d}
#| output: asis
d <- vuln_box(filter(vuln_tbl, stringr::str_detect(Code, "^D")))
tagList(!!!d)
```


## Version Note
This report was generated `r Sys.Date()` using version `r utils::packageVersion("ccviR")` of the ccviR package.



<!-- 
https://quarto.org/docs/authoring/includes.html#computations
Cannot create the {{< include _about.qmd >}} short code through an R block, 
because then the _about.qmd won't itself be executed (just copy/pasted, which is
fine if md, but not if qmd.

Fix from: https://stackoverflow.com/a/74301258 
-->
`r if (!include_about) "::: {.content-hidden}"`

:::{style="page-break-before: always"}
<!-- Add page break -->
:::
  
{{< include _about.qmd >}}

`r if (!include_about) ":::"`






