---
title: "Using the ccviR app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the ccviR app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
The ccviR app is an interactive application designed to offer a user-friendly and simple interface for using the [NatureServe Climate Change Vulnerability Index](https://www.natureserve.org/conservation-tools/climate-change-vulnerability-index) to identify species that are vulnerable to climate change. Effective use of the NatureServe CCVI tool requires some GIS analysis followed by entering the results of that analysis and answers to vulnerability questions into an Excel spreadsheet which then calculates the corresponding index. With this app we combine these two steps into one, using R to do the spatial analysis and calculate the index based on user inputs entered through a Shiny application. 

The ccviR app uses the ccviR package to calculate the NatureServe CCVI index based on user inputs but users familiar with R could use the ccviR package directly to investigate different aspects of the NatureServe CCVI algorithm. However, this vignette will only focus on using the app.

## Installation
The package can be installed by:  

1. Cloning the github [repo](https://github.com/see24/ccviR) to an RStudio project. [Instructions](https://happygitwithr.com/existing-github-first.html#new-rstudio-project-via-git-clone) (you only have to do part 16.2)
    i. If that doesn’t work save the R and inst folders in an RStudio project folder named ccviR, then go to step 2  
2. Running devtools::install() in the r console of that project


## Preparing the data
Before launching the ccviR app users will need to ensure that they have the appropriate data ready for the CCVI calculations. Climate exposure data has been provided with the app and can be downloaded from [Google Drive](https://drive.google.com/drive/folders/1YqtPpzzXvGVRI3BHg-yuKrhJL0rFnSGH?usp=sharing). This contains data for climate change exposure based on anticipated change in temperature and moisture, and the climate change exposure index (CCEI) for the non-breeding range; and sensitivity based on variation in temperature and precipitation. A shapefile can be added to this dataset to represent the physiological thermal niche of the species. 

In addition to climate data the user must supply polygon shapefiles of the species' range and the area being assessed. The user may optionally also provide a shapefile of non-breeding range for migratory species, and a habitat suitability raster showing change in the species range under climate change such that 1 represents habitat lost, 7 represents habitat gained, and 2-6 represent habitat maintained. 

This data should be saved under one directory with the climate data in a sub-directory that does not contain any other files. 

## Launching the app
The app is launched by calling `run_ccvi_app()` which will launch the app in the users default internet browser with file selection being relative to the current working directory. To run the app with data stored in a different folder you can call `run_ccvi_app()` with the folder path as the first argument. For example if all of the data is stored in a subdirectory named "data" of the RStudio project where I am calling the function I would call `run_ccvi_app("data")` to avoid having to open the data folder every time I select a file. In addition the ccviR package includes a demo data set that can be used to test out the app, to use this data set call `run_ccvi_app("demo")`. 

## Using the app
The first tab contains information about the species being assessed. Once all mandatory fields are filled in click next. The second tab is where the data will be loaded and displayed on an interactive map. All data in the climate data folder will be loaded and the names must match those in the demo dataset. Other files must be selected individually. Click Load to view the data on an interactive map and a table showing the exposure multiplier derived from the temperature and moisture data. The third tab includes all information on the species vulnerability to climate change and is based on expert knowledge and characteristics of the species. Users should use the NatureServe CCVI [guidelines](https://www.natureserve.org/sites/default/files/guidelines_natureserveclimatechangevulnerabilityindex_r3.02_1_jun_2016.pdf) to interpret each question. If the answer to a question is unknown leave the question blank. The next tab includes the vulnerability questions that are answered through spatial analysis of the data provided. For each question a map will show the data and a table will show the information extracted. The appropriate checkbox will be filled in based on the NatureServe guidelines, however a user may check or uncheck a box if they would like to adjust the value. If the data for the spatial analysis was not provided the map will be blank and the user may answer the question based on expert knowledge or leave it blank. The final tab shows the results of the climate change vulnerability index and some information about what factors had the most impact on the calculated index.  


```{r setup, eval=FALSE}
library(ccviR)
run_ccvi_app("demo")
```
