---
output:
  pdf_document: 
    keep_tex: true
    fig_caption: true
    # includes:
      #in_header: columns.tex # supposed to make 2 col layout but too advanced for me
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage[utf8]{inputenc}
  - \usepackage{makecell}
  - \usepackage{xcolor}
editor_options: 
  chunk_output_type: console
params: 
    out_data: !r list(scenario_name = c("RCP 4.5", "RCP 8.5"), index = c("EV", "HV"))
    clim_vars: !r list(mat = system.file("extdata/clim_files/processed/MAT_reclassRCP_4.5.tif", package = "ccviR"), cmd = system.file("extdata/clim_files/processed/CMD_reclassRCP_4.5.tif", package = "ccviR"))
    range_poly: !r system.file("extdata/rng_poly.shp", package = "ccviR")
    range_poly_clim: !r system.file("extdata/rng_poly.shp", package = "ccviR")
    scale_poly: !r system.file("extdata/assess_poly.shp", package = "ccviR")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE) 
out_data <- params$out_data
clim_vars <- params$clim_vars
range_poly <- params$range_poly
range_poly_clim <- params$range_poly_clim
scale_poly <- params$scale_poly

# should install TinyTex if not already installed
if(!tinytex::is_tinytex()){
  tinytex::install_tinytex()
}

library(purrr)
library(dplyr)
library(ccviR)
library(kableExtra)

# defined as latex colors below
cols1 <- c("#808080", "#008000", "#FFC125", "#FF8C00", "#FF0000")
cols <- c("gray", "green", "yellow", "orange", "red")
nms <- c("Insufficient Evidence", "Less Vulnerable", "Moderately Vulnerable",
         "Highly Vulnerable", "Extremely Vulnerable", "Not Applicable")
# function to color index value text
index_col <- function(ind){
  case_when(
    ind == "EV" ~ sprintf("\\textcolor{%s}{%s}", cols[5], nms[5]),
    ind == "HV" ~ sprintf("\\textcolor{%s}{%s}", cols[4], nms[4]),
    ind == "MV" ~ sprintf("\\textcolor{%s}{%s}", cols[3], nms[3]),
    ind == "LV" ~ sprintf("\\textcolor{%s}{%s}", cols[2], nms[2]),
    ind == "IE" ~ sprintf("\\textcolor{%s}{%s}", cols[1], nms[1]),
    ind == "N/A" ~ sprintf("\\textcolor{%s}{%s}", cols[1], nms[6])
    
  )
}

if(unique(out_data$common_name) == ""){
  ttl_name <- paste0("_", unique(out_data$species_name), "_")
} else {
  ttl_name <- paste0(unique(out_data$common_name),
                     " (_", unique(out_data$species_name), "_)")
}

if (identical(rmarkdown::metadata$runtime, "shiny")){
    shiny::setProgress(0)
}

```

<!-- make LaTeX colours that match our hex code colors -->
\definecolor{gray}{HTML}{808080}
\definecolor{green}{HTML}{008000}
\definecolor{yellow}{HTML}{FFC125}
\definecolor{orange}{HTML}{FF8C00}
\definecolor{red}{HTML}{FF0000}
---
title: `r ttl_name`
subtitle: Climate Change Vulnerability Index Report
---

## Climate Change Vulnerability Index
```{r results='asis'}
text <- paste0("### ", out_data$scenario_name, ": ", "\\LARGE", index_col(out_data$CCVI_index),
               collapse = "\n")
cat(text)
```


## Migratory Exposure Index
```{r results='asis'}
text <- paste0("### ", out_data$scenario_name, ": ", "\\Large",
               index_col(out_data$mig_exposure),
               collapse = "\n")
cat(text)
if (identical(rmarkdown::metadata$runtime, "shiny")){
    shiny::setProgress(0.1)
}
```


## Species range
```{r range_map,fig.height= 3, fig.width=7, fig.cap= "Map of the species range (red) and the assessment area (black)."}
# should add a context polygon that is North America to state/prov level as
# background then choose the map extent as whichever is bigger between range and assessment area
comb_ext <- sf::st_union(sf::st_bbox(range_poly_clim) %>% sf::st_as_sfc(),
                            sf::st_bbox(scale_poly) %>% sf::st_as_sfc() %>%
                              sf::st_transform(sf::st_crs(range_poly_clim)))

tmap::tm_shape(context_North_Am)+
  tmap::tm_polygons(col = "grey90", border.col = "grey80")+
  tmap::qtm(scale_poly, borders = "black", fill = NULL)+
  tmap::tm_shape(range_poly_clim, bbox = comb_ext, is.master = TRUE)+
  tmap::tm_borders(col = "red")+
  tmap::tm_add_legend(type = "line", labels = c("Range", "Assessment Area"), 
                      col = c("red","black"))+
  tmap::tm_layout(legend.outside = TRUE)
if (identical(rmarkdown::metadata$runtime, "shiny")){
  shiny::setProgress(0.3)
}
```


## Exposure 
```{r bivar_exp,fig.height= 3, fig.width=7, results='hide', fig.cap="Map of exposure based on both change in temperature and change in moisture regime."}
# plotly graphics don't show up in pdf. Either output to png or just print the text. 
# walk2(index_res$index, index_res$scenario_name, 
#       ~ccviR:::plot_index_gauge(.x, type = "", codes = c("IE","LV", "MV", "HV", "EV"),
#     nms = c("Insufficient Evidence",
#                "Less Vulnerable",
#                "Moderately Vulnerable",
#                "Highly Vulnerable",
#                "Extremely Vulnerable"),
#     cols = c("#808080", "#008000", "#FFC125", "#FF8C00",
#               "#FF0000"),
#     ttl = .y) %>% print()) 
bivar_exp_plot <- plot_bivar_exp(clim_vars$mat, clim_vars$cmd, scale_poly, range_poly)

gridExtra::grid.arrange(grobs = list(ggplot2::ggplotGrob(bivar_exp_plot$plot), 
                                     ggplot2::ggplotGrob(bivar_exp_plot$legend)), 
                        layout_matrix = rbind(c(1,1,1,1,1,2),
                                              c(1,1,1,1,1,2),
                                              c(1,1,1,1,1,2)))
```

```{r}
if (identical(rmarkdown::metadata$runtime, "shiny")){
    shiny::setProgress(0.7)
}
```



## Vulnerability Factors
```{r vuln_table, results='asis'}
coms <- out_data %>% 
  select(scenario_name, matches("com_[B,C,D]\\d.*")) %>% 
  tidyr::pivot_longer(contains("com"), names_to = "Code", values_to = "Comments",
                      names_prefix = "com_") 

valueNms <- c("Greatly increase", "Increase", "Somewhat increase", "Neutral", "Unknown")
names(valueNms) <- c(3, 2, 1, 0, -1)
cols2 <- cols %>% stats::setNames(rev(names(valueNms)))


vuln_qs <- out_data %>% 
  select(scenario_name, matches("^[B,C,D]\\d.*")) %>% 
  mutate(across(everything(), as.character)) %>% 
  tidyr::pivot_longer(-scenario_name, names_to = "Code", values_to = "Score") %>% 
  tidyr::separate(Score, into = paste0("Value", 1:4), fill = "right", sep = ", ",
                  convert = TRUE) %>%
  mutate(Value1 = tidyr::replace_na(Value1, -1)) %>% 
  mutate(score = 1) %>% 
  ccviR:::filter_applicable_qs(out_data$tax_grp, c("Vascular Plant", "Nonvascular Plant")) %>% 
  filter(!is.na(score)) %>% 
  rowwise() %>% 
  mutate(Score = c(.data$Value1, .data$Value2, .data$Value3, .data$Value4) %>% 
           stats::na.omit() %>% range() %>%  
           unique() %>% as.character() %>% 
           {text_spec(valueNms[.], format = "latex", background = cols2[.], bold = TRUE)} %>%
           paste0(collapse = " - "))


vuln_tbl <- left_join(vuln_qs, coms, by = join_by(scenario_name, Code)) %>% 
  left_join(vulnq_code_lu_tbl,by = join_by(Code)) %>% 
  select(scenario_name, Question, Code, Score, Comments)

# Table version but might be weird if long comments which are likely and recommended
# vuln_tbl %>% nest_by(scenario_name) %>% pull(data) %>% 
#   purrr::walk2(out_data$scenario_name,
#                ~print(kbl(.x, digits = 3, booktabs = FALSE, longtable = TRUE,
#                           label = .y, escape = FALSE,
#                           format = "latex", caption = .y, linesep = "") %>% 
#                         column_spec(1, "20em") %>%
#                         column_spec(4, "20em") %>%
#                         kable_styling(latex_options = c("repeat_header"))))

# paragraph version 
vuln_tbl %>% nest_by(scenario_name) %>% pull(data) %>%   
  purrr::walk2(out_data$scenario_name,
               \(df, scn_nm){
                 cat(paste0("\n\n### ", scn_nm, "\n\n"), 
                     paste0("#### ", df$Code, ": ", df$Question, "\n",
                            "\\hspace{-2.5em} **Score: **", df$Score, "\\newline \n", 
                            "**Comments: **", df$Comments, collapse = "\n\n"), sep = "")
               })

```

```{r}
if (identical(rmarkdown::metadata$runtime, "shiny")){
    shiny::setProgress(0.9)
}
```

### Version Note
This report was generated with version `r utils::packageVersion("ccviR")` of the ccviR package.
